---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data
```{r , message=FALSE}
library(readr)
library(lubridate)

activity <- read_csv("activity.csv")
activity$date <- ymd(activity$date)
activity$day <-weekdays(as.Date(activity$date))

summary(activity)

```



## What is mean total number of steps taken per day?

```{r, cache=TRUE, message=FALSE, warning=FALSE}

library(tidyverse)

# Calculating total steps per day
steps_per_day <- activity %>% 
        group_by(date) %>% 
        summarise(total_steps = sum(steps, na.rm = TRUE))

colnames(steps_per_day) <- c("Date", "Steps")

print(steps_per_day)
```

Creating histogram of total number of steps taken each day

```{r total steps}
hist(steps_per_day$Steps, breaks = 5, xlab = "Steps", main = "Total Steps Per Day")
```


```{r}
mean_steps <- mean(steps_per_day$Steps, na.rm = TRUE)

median_steps <- median(steps_per_day$Steps, na.rm = TRUE)
```

The mean and median number of steps taken per day are `r mean_steps` and `r median_steps` respectively.


## What is the average daily activity pattern?

```{r daily pattern, warning=F, message=F}
library(tidyverse)

daily_pattern <- activity %>% 
        group_by(interval) %>% 
        summarise(average = mean(steps, na.rm = TRUE))


p <- ggplot(daily_pattern, aes(x=interval, y=average), xlab = "Interval", ylab="Average Number of Steps")

p + geom_line()+xlab("Interval")+ylab("Average Number of Steps")+ggtitle("Average Number of Steps per Interval")

```


```{r}
max_steps_interval <- which.max(daily_pattern$average)

max_steps <- daily_pattern$interval[max_steps_interval]
```

The 5-minute interval which had the maximum number of steps is `r max_steps` interval.



## Imputing missing values

```{r}
missing <- sum(is.na(activity$steps))
```

Total number of rows with missing values is `r missing`.


The strategy for filling missing values will be to substitute missing steps with average steps for that interval.

```{r, message=FALSE, cache=TRUE}


library(tidyverse)

StepsPerInterval <- tapply(activity$steps, activity$interval, mean, na.rm = TRUE)
# split activity data by interval
activity_split <- split(activity, activity$interval)
# fill in missing data for each interval
for(i in 1:length(activity_split)){
    activity_split[[i]]$steps[is.na(activity_split[[i]]$steps)] <- StepsPerInterval[i]
}
activity_imputed <- do.call("rbind", activity_split)
activity_imputed <- activity_imputed[order(activity_imputed$date) ,]

activity_imputed$steps <- ceiling(activity_imputed$steps)

```


  
First ten rows of new data set is shown below:


```{r, warning=F}
head(activity_imputed, 10)

```

Creating Histogram for total number of steps per day

```{r total steps imputed, message=F, warning=F}
library(tidyverse, quietly = TRUE)
total_steps_per_day <- activity_imputed %>% 
        group_by(date) %>% 
        summarise(total_steps = sum(steps))

colnames(total_steps_per_day) <- c("Date", "Steps")

hist(total_steps_per_day$Steps, breaks = 5, xlab = "Steps", main = "Total Steps Per Day")
```


```{r}
mean_step <- mean(total_steps_per_day$Steps, na.rm = TRUE)

median_step <- median(total_steps_per_day$Steps, na.rm = TRUE)
```

The mean and median number of steps taken per day are `r mean_step` and `r median_step` respectively.


The mean number of daily steps has changed a little but median remains almost same.
However, the overall shape of the distribution has not changed.

## Are there differences in activity patterns between weekdays and weekends?


```{r}
activity_imputed$day_type <- ifelse(activity_imputed$day == "Saturday"|activity_imputed$day == "Sunday", "weekend", "weekday")

activity_imputed$day_type <- as.factor(activity_imputed$day_type)
```

```{r steps per time for different weekdays}
# create table with steps per time across weekday days or weekend days

StepsPerTimeDT <- aggregate(steps ~ interval + day_type, data=activity_imputed, FUN=mean, na.action=na.omit)



# draw the line plot
j <- ggplot(StepsPerTimeDT, aes(interval, steps))
j+geom_line(col="darkred") + ggtitle("Average steps per time interval: weekdays vs. weekends") + xlab("Interval") + ylab("Steps") + theme(plot.title = element_text(face="bold", size=12)) + facet_grid(day_type ~ .)


```


The step activity trends are different based on whether the day occurs on a weekend or not. This may be due to people having an increased opportunity for activity beyond normal work hours for those who work during the week.
